---
title: "Wearable Device Effect on Personal Health"
author: "Chris Brofft"
date: "7/8/2018"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Course Project Instruction

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity fairly easily.
Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification(A) throwing the elbows to the front(B), lifting the dumbbell only halfway(C), lowering the dumbbell only halfway(D), and throwing the hips to the front(E) .

The training data for this project are available here:  
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv  
The test data are available here:  
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv  
The data for this project come from this source:   http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.  

*The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.*


## Getting and cleaning data

Load data and packages

```{r getting, message=FALSE, warning=FALSE}
TrainData <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
TestData <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
train <- read.csv(TrainData)
test <- read.csv(TestData)
library(dplyr)
library(caret)
```

Number of records per classe

```{r exploring, message=FALSE, warning=FALSE}
summarize(group_by(train, classe), Training=n())
```

  
Find columns with lot NA or blank values

```{r finding, message=FALSE, warning=FALSE}
na <-sapply(train, function(y) sum(length(which(is.na(y)))))
na <- data.frame(na)
na <- tibble::rownames_to_column(na)
blank <- sapply(train, function(y) sum(length(which(y==""))))
blank <- data.frame(blank)
blank <- tibble::rownames_to_column(blank)
na_blank <- full_join(na, blank, by=c("rowname","rowname"))
na_blank <- mutate(na_blank, NaBlank=na+blank, NaBlankPerc=(na+blank)/nrow(train)*100)
filter(na_blank, NaBlank > 0)
```

100 of 160 variable are missing for 98% of the values. 

```{r cleaning, message=FALSE, warning=FALSE}
not_na <- filter(na_blank, NaBlank == 0)
Col <- not_na$rowname
train <- select(train,one_of(Col))
train = train[-c(1,3,4,5,6,7)]
```

Columns used for model:

```{r finalData, message=FALSE, warning=FALSE}
str(train)
```


## Split data

I will be spliting the data into 2 groups: one that has 75% of cases, and the other with 25%.

```{r training, message=FALSE, warning=FALSE}
inTrain = createDataPartition(train$classe, p = 0.75, list=FALSE)
MyTrain = train[inTrain,]
MyTest = train[-inTrain,]
```

I will now use the training data in building the model


## Models

### Classification Tree

First, a Classification Tree model:

```{r modelTree, message=FALSE, warning=FALSE}
set.seed(10000)
modelTree <- train(classe~., data = MyTrain[-1], method="rpart")
modelTree
```

The accuracy is 51%.  
Next I test the model.  

```{r predictionTree, message=FALSE, warning=FALSE}
predTree <- predict(modelTree, newdata = MyTest)
table(predTree,MyTest$classe)
TruePredictionTree <- data.frame(classe=MyTest$classe,esito=predTree==MyTest$classe)
frequenzeTree <- table(TruePredictionTree$classe,TruePredictionTree$esito)
frequenzeTree
prop.table(frequenzeTree,1)*100
colSums(frequenzeTree)
colSums(frequenzeTree)/sum(frequenzeTree)
```

Only about 50% true. Very bad in particular for classe D.


### Random Forest

I now try with a Random Forest model.

```{r modelRandomForest, message=FALSE, warning=FALSE}
set.seed(10000)
modelRF <- train(classe~., data = MyTrain[-1], method="rf")
modelRF
```

Testing the random forest model:

```{r predictionRandomForest, message=FALSE, warning=FALSE}
predRF <- predict(modelRF, newdata = MyTest)
table(predRF,MyTest$classe)
TruePredictionRF <- data.frame(classe=MyTest$classe,esito=predRF==MyTest$classe)
frequenzeRF <- table(TruePredictionRF$classe,TruePredictionRF$esito)
frequenzeRF
prop.table(frequenzeRF,1)*100
colSums(frequenzeRF)
colSums(frequenzeRF)/sum(frequenzeRF)
```

Random forest gaves a much better prediction, 99,4% true. I am choosing this model for the final predictions.


## Predictions

I make the prediction using the Random Forest model on the 20 cases provided.

```{r TestPrediction, message=FALSE, warning=FALSE}
predict(modelRF, newdata = test)
20*0.994
```

I expect to have only 0.6% of error, so 20 on 20 or at least 19 on 20 are correct predictions.